*cellcount*: an open-source and automated approach to cyanobacteria enumeration

### 2023-02-24 draft_manuscript.Rmd
Updated.

Several methods have been established to understand HAB dynamics, primarily to develop monitoring protocols and establish mitigation 
practices. Microscopic enumeration has been considered as the gold standard for several decades as the method is widely available and simple to use [@Jin2018]. Although it is the standard method, microscopic enumeration can be time consuming, can be limited to small sample quantities, and requires significant knowledge of species identification [@Li2014] and [@Jin2018]. Technology has become significant in advancing scientific methods, and it is no less significant in its role in advancing enumeration processes. Several commercially available products such as the FlowCam (Fluid Imaging Technologies Inc.) and Imaging FlowCytobot (McLane Research Laboratories Inc.) can be used to simplify enumeration methods (REF). These technologies can be accurate, demonstrate time-savings, and simplify data collection (REF). However, several limitations can be observed such as particle size, misidentification of cells, and overall grouping of particles within samples, which can require assistance from researchers (REF). In addition, these technologies are expensive to purchase and maintain, reducing the nature of these technologies being widely available [@Stauffer2019]. A few studies have explored the process of automated cell enumeration methods by means of technology and software development, but promoting widespread availability. ImageJ is an open-source imaging analysis software and has available software for particle/cell enumeration [@Schulze2011]. Other software algorithms have been developed to accomplish automated cellular enumeration [@Jin2018]. These methods can still be considered time-consuming (ImageJ) and require extensive knowledge of mathematical modeling [@Jin2018].

The methods mentioned previously can be viable options for researchers regarding monitoring purposes, but can be expensive and require 
extensive training and materials. This leads to the suggestion of the need for accurate, low-cost, and widely available methods for rapid and simplistic automated enumeration of cells. A report released by [@Pokrzywinski2019] outlines an automated cyanobacteria enumeration script using the open-source programming language, R. [@Pokrzywinski2019] shows that while using fluorescent microscopy images, this enumeration script performs considerably well compared to standard microscopic enumeration methods. However, the previous presentation of this script can be considered extremely difficult to grasp for inexperienced R users. We utilized this script as a template for the development of an R package, involving the reduction of code into simple functions, expanding the ability for end-user modifications, development of a vignette for inexperienced users, and promoting open-source availability.

@@

*analyze_images* uses a simplistic Shiny UI which consists of two buttons, one to run the analysis and the other to close the UI interface.

@@

@@

@@

>>>>

Several methods have been established to understand HAB dynamics, primarily to develop monitoring protocols and establish mitigation 
practices. Microscopic enumeration has been considered as the gold standard for several decades as the method is widely available and simple to use [@Jin2018]. Although it is the standard method, microscopic enumeration can be time consuming, can be limited to small sample quantities, and requires significant knowledge of species identification [@Li2014] and [@Jin2018]. Technology has become significant in advancing scientific methods, and it is no less significant in its role in advancing enumeration processes. Several commercially available products such as the FlowCam (Fluid Imaging Technologies Inc.) and Imaging FlowCytobot (McLane Research Laboratories Inc.) can be used to simplify enumeration methods (REF). **These technologies can be accurate, demonstrate time-savings, and simplify data collection (REF). However, several limitations can be observed such as particle size, misidentification of cells, and overall grouping of particles within samples, which can require assistance from researchers (REF).** In addition, these technologies are expensive to purchase and maintain, reducing the nature of these technologies being widely available [@Stauffer2019]. A few studies have explored the process of automated cell enumeration methods by means of technology and software development, but promoting widespread availability. ImageJ is an open-source imaging analysis software and has available tools for particle/cell enumeration [@Schulze2011]. Other software algorithms have been developed to accomplish automated cellular enumeration [@Jin2018]. These methods can still be considered time-consuming (ImageJ) and require extensive knowledge of mathematical modeling [@Jin2018].

The methods mentioned previously can be viable options for researchers regarding monitoring purposes, but can be expensive and require 
extensive training and materials. This leads to the suggestion of the need for accurate, low-cost, and widely available methods for rapid and simplistic automated enumeration of cells. A report released by [@Pokrzywinski2019] outlines an automated cyanobacteria enumeration script using the open-source programming language, R. [@Pokrzywinski2019] shows that while using fluorescent microscopy images, this enumeration script performs considerably well compared to standard microscopic enumeration methods. However, the previous presentation of this script can be considered extremely difficult to grasp for inexperienced R users and lacks organization. We utilized this script as a template for the development of an R package, involving the reduction of code into simple functions, expanding the ability for end-user modifications, development of a vignette for inexperienced users, and promoting open-source availability.

@@

*analyze_images* uses a simplistic Shiny UI which consists of two buttons, one to run the analysis and the other to close the UI interface. Once the proper data has been entered from the *input_data* function, users simply click the "Run Image Analysis" button to begin. Sounds from the ***beepr*** package alert users of completion of the analysis, where the "Close window" button exits the UI window. Users can navigate to their specific directories to find analyzed images and related .csv files for their review.

(INSERT IMAGE OF *analyze_images* SHINY APP HERE)

@@

The current version of this package may be limited for some users regarding processing power. Analysis with this package uses local memory, which may be limiting if expected to analyze large amounts of images in batches. We recommend at least 16GB of localized RAM, the use of a cloud computing supercomputer, or organize images in easier-to-analyze batches to reduce local memory strain.

Users must be aware that this package does not contain abilities to identify distinct cells. The development of this package utilized carefully cultured monoculture of various cyanobacteria species. Users may find difficulty in using this analysis tool if cultures contain heavy amounts of particulates, other cellular debris, and contamination of other species. It is advised to use this tool for single-species only. Future developments of this tool include identifying features and building image libraries of specific cellular species for use in environmental samples.

Outside of package limitations, users must be aware of limitations that may arise from poor microscopy imaging; reduce lighting as much as possible, avoid heavy background fluorescence, and so forth. Good images produce good results - keep this in mind as you utilize this tool.

@@

***cellcount*** provides an open-source approach to cyanobacteria enumeration that is accurate and rapid compared to standardized microscopy methods. The open-source approach allows researchers to access this tool for free and to modify and make additions the existing code to fulfill their particular needs, whereas the Shiny GUI allows users of all coding backgrounds to utilize this analysis tool.

@@

We acknowledge...
